\Wxxsir methods to estimate $\rr$ in the SIR model with applications to 2009 pandemic influenza

### 2018-09-06 paper/Version16/draft_v16.tex
Changed images to jpg
pulling
Added new explanations and comments

\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
%\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}% http://ctan.org/pkg/graphicx
\usepackage{systeme}
%% tables
\usepackage{booktabs}
\usepackage{lscape}
\usepackage[table,xcdraw]{xcolor}
\usepackage{float}
\usepackage{url}

%% Make red text
\newcommand{\com}[1]{\textcolor{red}{ #1}}

@@

@@

Denote $\hat{X}(t), \hat{Y}(t)$, and $\hat{Z}(t)$ to be the observed number in each compartment and $X(t), Y(t)$, and $Z(t)$ to be the true, underlying model at time $t$.   The stochastic model is given by
\begin{align*}
  \hat{X}(t+1) &= \hat{X}(t) - x_t\\
  \hat{Y}(t+1) & = N - \hat{X}(t+1) - \hat{Z}(t+1), \nonumber \\
  \hat{Z}(t+1) &= \hat{Z}(t) + z_t \nonumber
\end{align*}
with $\hat{X}(0)$, $\hat{Y}(0)$, and $\hat{Z}(0)$ known. The random terms in Equation \eqref{eq:sir-noise} are random variables.  Here we choose them to have the following distributions,
\begin{align*}
  x_{t} &\sim \text{Binomial}\left (\hat{X}(t), \frac{\beta Y(t)}{N} \right ) \\
  z_{t} &\sim \text{Binomial}\left (\hat{Y}(t), \gamma \right ). \nonumber
\end{align*}
Then the likelihood is
\begin{align*}
  \mathcal{L}\left (\beta, \gamma ; \textnormal{Data} \right ) &=
                                                  \prod_{t=0}^{T-1} \left [\left ( \begin{array}{c}
                                                            \hat{X}(t) \\
                                                            x_t
                                                            \end{array}
  \right ) \left (\beta - \frac{\beta\left (X(t) +Z(t) \right )}{N} \right )^{x_t} \left (1 - \beta +  \frac{\beta\left (X(t) + Z(t) \right )}{N} \right )^{\hat{X}(t) - x_t} \cdot \right .\\
  & \left . \left ( \begin{array}{c}
              N -\left ( \hat{X}(t) + \hat{Z}(t)\right ) \\
              z_t
            \end{array}
  \right ) \left (\gamma \right )^{z_t} \left ( 1 - \gamma\right )^{N -\left ( \hat{X}(t) + \hat{Z}(t)\right ) - z_t} \right ].
\end{align*}
The estimate for $\rr$ is then
\begin{align}\label{eq:lbe-r0}
  (\hat{\beta}, \hat{\gamma}) &= \arg \max_{\beta, \gamma} \mathcal{L}\left ( \beta, \gamma; \textnormal{Data} \right )\nonumber\\
  \rr &= \frac{\hat{\beta}}{\hat{\gamma}}
  \end{align}
  The two main differences between this model and MC are  that 1) LBE looks at the joint likelihood of the XYZ model, whereas MC only explicitly models the number of susceptibles and 2) the difference in the structure of the transition probability.  We use the delta method in conjunction with $\hat{\beta}$ and $\hat{\gamma}$ estimations from the Nelder-Mead optimization of the likelihood to estimate the stantard error of $\hat{\rr}$.
  The advantages of using this model are that our modelling assumptions are explicit and that the modelling process seeems to be reasonable.  Disadvantages include maximizing the liklihood may be more time consuming than other methods due to the need to approximate $X(t)$ and $Z(t)$ for many different values of $\beta$ and $\gamma$ and that if we are wrong about the generating process of the XYZ curves, then this model may be a very poor fit.

@@

We can also use the Delta method to derive variance estimates of LMA, LMAT, LL, (which in this case is equivalent to the standard error derived from the typical linear regression framework).  We can also use the Delta method to derive an estimate for MC, where $\hat{\sigma}$ is the estimate of the variance of the $\hat{\alpha}$,
\begin{align*}
h(\alpha) &= \log \left ( \frac{1}{1-\alpha}\right ) \\
  h^\prime(\alpha) &= \alpha - 1 \\
                     V[ \hat{\rr} ] & \approx \frac{1}{T}  \left (\frac{\hat{\sigma}}{(\hat{\alpha} - 1)^2} \right ).
\end{align*}

@@

      \end{center}
      The baseline data set is generated by adding noise to the susceptible and infectious curves from deterministic SIR paths, which are shown in Figure \ref{fig:baseline-data}.  The recovered compartment is then adjusted to maintain a constant population value.  In this scenario, we see that the infection begins slowly but over the course of a year infects nearly 75\% of the population.
      \begin{figure}
        \centering
        \includegraphics[width=.8\textwidth]{images/baseline_sir.pdf}
        \caption{SIR (XYZ) curves for the deterministic model with $N=100,000$, $(\beta, \gamma) = (0.06, 0.03)$, $(X(0), Y(0), Z(0))= (99950, 50,0)$  The baseline data is generated by using this model and adding noise to the $X$ and $Y$ compartments and adjusting $Z$ at each time step to maintain a constant population.}\label{fig:baseline-data}
        \end{figure}
      This baseline data set is chosen due to its moderate $\rr$ size of 2, which indicates an outbreak of an epidemic similar to HIV, ZIK-V, EVD, and H1N1 pandemic influenza (see Table \ref{tab:r0-real-ex}).  The infection rate and recovery rate ($\beta, \gamma$) are relatively small.  The duration of the disease, under the true model, spans nearly a whole year (see Figure \ref{fig:baseline-data}) and so we use a year's worth of daily observations of ($\hat{X},\hat{Y}, \hat{Z}$) values.  Initially, 50 (0.05\%) infectious individuals are introduced to the population of 100,000 individuals (approximately the average population size of a US county in 2015).  The standard deviation of the $\hat{X}$ values is 0.1\% of the magnitude of the initial number of susceptible individuals, and the standard deviation of the $\hat{Y}$ values is 10\% of the magnitude of the initial number of infectious individuals, although in this setting, the number of infectious individuals increases rapidly.  We use Gaussian error as our baseline error because of its well known statistical properties so we can more easily compare methods to one another.  Later on, we explore the autoregressive and PAVA simulated data, which may better simulate data from  a real epidemic.

@@

\section{Results}\label{sec:results-top}
\subsection{$\rr$ estimation methods}\label{sec:results}
There are numerous aspects to consider when using a method to estimate $\rr$.  Among these include values of both the infection and recovery rate $(\beta, \gamma)$, the number of data points available ($T$), the total size of the population ($N$), the initial percent of susceptible individuals $\left (\frac{X(0)}{N}\right)$, the initial percent of infectious individuals $\left (\frac{Y(0)}{N}\right )$, the magnitude of the variance of both the number of susceptibles and the number of infectious ($\sigma_X^2, \sigma_Y^2$), the error modeling assumptions, data generated from a non-SIR model, and finally methods of estimating the variance.  For the sake of brevity, we do not provide the full results here, and instead examine the results at a high-level.  However, for those interested readers, we provide results in the form of forest plots and tables of estimates of $\rr$ for each data set in the Supplementary Material.

We implement the \wxxsir methods to estimate $\rr$ using the baseline data set with normal error.  The results are displayed in Figure \ref{fig:baseline-res} and Table \ref{tab:baseline-res}.
\begin{figure}[H]
  \centering
  \includegraphics[scale=0.5]{images/BaseBase.tiff}
  \caption{Forest plots of estimates of $\rr$ for the \xxsir methods for the baseline data set $(\beta=.06, \gamma=.03, O=365, X(0)=99950, Y(0)=50, Z(0)=0, \sigma_X=100, \sigma_Y=5, N=10^5)$.  The points are the estimates of $\rr$ and the lines denote $\pm 2\cdot $Std. Err.  The vertical, dashed line is the true $\rr$ value.}\label{fig:baseline-res}
  \end{figure}

\begin{table}[H]	
	\centering
	\begin{tabular}[t]{l|r|r}
		\hline
		Model & Estimate & Std. Err\\
		\hline
		RE & 2.0000 & 0.0056\\
		\hline
		rRE & 1.9997 & 0.0050\\
		\hline
		LMA & 1.6864 & 0.1886\\
		\hline
		LMAT & 1.4063 & 0.6309\\
		\hline
		IPR & 4.4204 & 12.3593\\
		\hline
		SIPR & 1.8871 & $<$ 1e-04 \\
		\hline
		LL & 1.9998 & 0.0002\\
		\hline
		MC & 0.9486 &  $<$ 1e-04 \\
		\hline
		SB & 1.6495 & 0.0672\\
		\hline
		LBE & 14.6784 & 1.2330\\
		\hline
	\end{tabular}
        \caption{Baseline Data $(\beta=.06, \gamma=.03, T=365, X(0)=99950, Y(0)=50, Z(0)=0, \sigma_X=100, \sigma_Y=5, N=10^5)$, $\rr$ Estimates and Standard Errors}\label{tab:baseline-res}
\end{table}

Overall, in Figure \ref{fig:baseline-res}, we see a wide variation in the estimates of and the standard errors for $\rr$ depending on the method used.  The methods of RE, rRE, LMA, LMAT, IPR, and LL all result in a CI that covers the true value of $\rr=2$.  Of these, LMAT and IPR result in a CI which contains 1, so we would not be able to say whether or the outbreak of a disease was due to random chance or not. On the other hand, many methods appear to underestimate the true $\rr$, and many of them have larger standard errors as well.  We would say that IPR's CI is uninformative. Additionally, MC and SIPR models seem to have unreasonably small standard errors.

@@

IPR often produces reasonable point estimates for $\rr$ but simply has too large of a CI to be considered a useful estimate.  On the other hand, SIPR has a small CI that typically underestimates the true value of $\rr$.  Since SIPR smoothes the SIR curves, the peak of the infectious curve in particular is shrunken.  Since the peak value of number of infectious is crucial in estimating $\rr$ for SIPR, it makes sense that SIPR underestimates this result.  Additionally, LMA and LMAT are generally prone to underestimating $\rr$ and have large and generally uninformative CIs.

@@

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.5]{images/start_arm.tiff}
	\caption{Forest plots of estimates of $\rr$ for the \xxsir methods for the Inits1-4 data sets $(\beta=.06, \gamma=.03, T=365,  Z(0)=0, \sigma_X=100, \sigma_Y=5, N=10^5)$.  The points are the estimates of $\rr$ and the lines denote $\pm 2\cdot $Std. Err.  The vertical line is the true $\rr$ value.}\label{fig:ar-small-I}
\end{figure}

One interesting feature we see in the results is that RE and rRE can result in very different estimations and CIs, which may seem unintuitive since rRE is a simple reparametrization of RE.  For example, in Figure \ref{fig:ar-r0}, we see that when the true value of $\rr$ is 1, rRE has a much smaller CI than RE, which may be one reason to prefer rRE to RE.  In this figure, we see when $\gamma << \beta$ (e.g. $\rr=60)$, then all estimates for $\hat{\rr}$ fail.  However, when $\beta >. \gamma$, we the estimates (besides IPR) can accurately obtain the true estimate. This figure also helps to show that RE and rRE are robust to extreme values of $\beta$ and $\gamma$ and hence extreme values of $\rr$.  \com{(For R0=1.5, why do RE, rRE, and LL look like 2 and not 1.5?))}

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.5]{images/parchange_ar.tiff}
	\caption{Forest plots of estimates of $\rr$ for the \xxsir methods for the infrec1-4 data sets $(T=365, X(0)=99950, Y(0)=50, Z(0)=0, \sigma_X=100, \sigma_Y=5, N=10^5)$.  The points are the estimates of $\rr$ and the lines denote $\pm 2\cdot $Std. Err.  The vertical line is the true $\rr$ value.}\label{fig:ar-r0}
\end{figure}

@@

%% model misspecification is terribly unrobust
We also examine circumstances where the data is generated from underlying models different from the SIR model.  The results are plotted in Figure \ref{fig:mod-rob}  All three sets of data sets generated by the non-SIR models result in poor estimates of $\rr$, meaning that even our best estimation techniques fail under model misspecification.  Moreover, these methods seem to underestimate $\rr$ \com{Illegible plot, Does RR=0?}

@@

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.5]{images/timediff.tiff}
	\caption{Forest plots of estimates of $\rr$ for the \xxsir methods for the baseline data sets with normal error structure $(\beta=.06, \gamma=.03, X(0) = 99950, Y(0)=50, \sigma_X=100, \sigma_Y=5, N=10^5)$.  The points are the estimates of $\rr$ and the lines denote $\pm 2\cdot $Std. Err.  The vertical line is the true $\rr$ value.}\label{fig:timediff}
      \end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.5]{images/obs.tiff}
	\caption{Forest plots of estimates of $\rr$ for the \xxsir methods for the baseline data sets with normal error structure $(\beta=.06, \gamma=.03, X(0) = 99950, Y(0)=50, \sigma_X=100, \sigma_Y=5, N=10^5)$.  The points are the estimates of $\rr$ and the lines denote $\pm 2\cdot $Std. Err.  The vertical line is the true $\rr$ value.}\label{fig:obs}
\end{figure}

@@

Table \ref{tab:rep-samp} shows the mean and standard deviation of $\rr$ estimates produced by the models under 100 repetitions of the baseline conditions, with new noise generated each time. We want to compare the standard deviations of the estimates to the standard errors returned by each of the models for a single data set to asses how reliable those standard errors are. Overall, it appears than most of the models have standard error estimates that are larger than the actual standard deviation of the estimates under repeated sampling. In some cases, such as LL, LMA, and LMAT, the standard deviation and standard error estimates are fairly close. The least squares methods, RE and rRE, on the other hand, tend to overestimate the error by an order of magnitude. Finally, SIPR and SB, have individual standard error estimates that are smaller than the standard deviation of repeated estimates, meaning that they understate how much error is in the estimates.

@@

Table \ref{tab:bb-samp} shows the mean and standard deviation of $\rr$ estimates produced by the models under 100 block bootstrap repetitions of the baseline data. The data were first detrended using the predictions of the compartment sizes at each time point from each of the models on the original data set. The block bootstrap algorithm was then applied on to the residuals from the predictions using a fixed block length. Only 5 of the models provide predictions of the size of the compartments and thus allow for detrending of the data. The ratio estimator and reparameterized model have a fairly accurate estimates of $\rr$ as well as very small standard errors. On the other hand, the linear approximation and smoothed IPR models both have very inaccurate estimates as well as very large standard errors of $\rr$.

@@

Figure \ref{fig:coverage} shows the proportion of repetitions for both repeated sampling and block bootstrapping in which the 95\% confidence intervals of the estimates produced by the models covers the true $\rr$. The ratio estimator, reparameterized, and linear model approximation with all time points contain the true $\rr$ value in all of the confidence intervals. On the other hand, the smoothed IPR, Markov chain, and sequential Bayes models never cover the true $\rr$ in any of the confidence intervals.

\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.5]{images/coverage2.tiff}
		\caption{Proportion of coverage of true $\rr$ of confidence intervals produced by repeated sampling for the applicable method (delta method or posterior).}
		\label{fig:coverage}	
	\end{center}
\end{figure}

\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.5]{images/coverage.tiff}
		\caption{Proportion of coverage of true $\rr$ of confidence intervals produced by repeated sampling for block bootstrapping.}
		\label{fig:coverage}	
	\end{center}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{images/h1n1.tiff}
	\caption{Results from SIR models on H1N1 data.}
\end{figure}

@@

\com{Results TBD}

@@

%% what is r0 and why it is important
In this paper, we examine $\rr$, the average number of secondary infections a primary individual generates when introduced into a completely susceptible population, in the context of the SIR model.  The quantity $\rr$ summarizes an epidemic into a single estimate in the hopes that infectious disease outbreaks may be compared to one another.  A value of $\rr > 1$ indicates an outbreak of a disease whereas a value of $\rr < 1$ indicates that no outbreak will occur.  Thus, it is critical to not only estimate the value of $\rr$ but also a corresponding CI.  Overestimating $\rr > 1$ during an ongoing epidemic may result in a waste of resources in prevention and control measures; underestimating $\rr < 1$ may result in a worldwide crisis.

In Table \ref{tab:r0-real-ex}, we provide estimates for $\rr$ that have appeared over the past three decades for HIV, Zika, Ebola, influenza, H1N1 influenza, and measles in different locations throughout the world.  Each of these estimates is greater than $1$, which makes sense as we tend to care more when a disease has the potential outbreak.  The largest $\rr$ point estimate we have seen is for measles at $\rr=17$ in England in the 20th century, and the smallest is $\rr=1.08$ for HIV in Portugal in the 21st century.
%% r0 as a property of the model
The quantity $\rr$ may summarize an epidemic, accounting for temperature, seasonality, location, and genetics of a disease, but it is still a property of the model.  Namely, how individuals progress through an epidemic (e.g. susceptible to infectious to recovered compartments) influences how we view $\rr$.  We thus should compare estimates for $\rr$ for a fixed epidemic model.  Here, we limit our analysis to estimates of $\rr$ derived from the SIR model.  Then $\rr = \frac{\beta}{\gamma}$ where $\beta$ is the rate of infection and $\gamma$ is the rate of recovery.  As a consequence, we need to estimate $\rr$ from not only the number of infections but also the number of individuals recovering at each time step.

>>>>

\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
%\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}% http://ctan.org/pkg/graphicx
\usepackage{systeme}
%% tables
\usepackage{booktabs}
\usepackage{lscape}
\usepackage[table,xcdraw]{xcolor}
\usepackage{float}
\usepackage{url}
%% Make red text
\newcommand{\com}[1]{\textcolor{red}{ #1}}

@@

%% tif -> png
\usepackage{epstopdf}
\epstopdfDeclareGraphicsRule{.tiff}{png}{.png}{convert #1 \OutputFile}
\AppendGraphicsExtensions{.tiff}

@@

Denote $\hat{X}(t), \hat{Y}(t)$, and $\hat{Z}(t)$ to be the observed number in each compartment and $X(t), Y(t)$, and $Z(t)$ to be the true, underlying model at time $t$.   The stochastic model is given by
\begin{align*}
  \hat{X}(t+1) &= \hat{X}(t) - x_t\\
  \hat{Y}(t+1) & = N - \hat{X}(t+1) - \hat{Z}(t+1), \nonumber \\
  \hat{Z}(t+1) &= \hat{Z}(t) + z_t \nonumber
\end{align*}
with $\hat{X}(0)$, $\hat{Y}(0)$, and $\hat{Z}(0)$ known. The terms $x_t$ and $z_t$ in Equation \eqref{eq:sir-noise} incorporate random behavior into the model.  Here we choose them to have the following distributions,
\begin{align*}
  x_{t} &\sim \text{Binomial}\left (\hat{X}(t), \frac{\beta Y(t)}{N} \right ) \\
  z_{t} &\sim \text{Binomial}\left (\hat{Y}(t), \gamma \right ). \nonumber
\end{align*}
Then the likelihood is, when we assuming that the next time step is dependent only on the current time step,
\begin{align*}
  \mathcal{L}\left (\beta, \gamma ; \textnormal{Data} \right ) &=
                                                  \prod_{t=0}^{T-1} \left [\left ( \begin{array}{c}
                                                            \hat{X}(t) \\
                                                            x_t
                                                            \end{array}
  \right ) \left (\beta - \frac{\beta\left (X(t) +Z(t) \right )}{N} \right )^{x_t} \left (1 - \beta +  \frac{\beta\left (X(t) + Z(t) \right )}{N} \right )^{\hat{X}(t) - x_t} \cdot \right .\\
  & \left . \left ( \begin{array}{c}
              N -\left ( \hat{X}(t) + \hat{Z}(t)\right ) \\
              z_t
            \end{array}
  \right ) \left (\gamma \right )^{z_t} \left ( 1 - \gamma\right )^{N -\left ( \hat{X}(t) + \hat{Z}(t)\right ) - z_t} \right ].
\end{align*}
The estimate for $\rr$ is then
\begin{align}\label{eq:lbe-r0}
  (\hat{\beta}, \hat{\gamma}) &= \arg \max_{\beta, \gamma} \mathcal{L}\left ( \beta, \gamma; \textnormal{Data} \right )\nonumber\\
  \rr &= \frac{\hat{\beta}}{\hat{\gamma}}
  \end{align}
  The two main differences between this model and MC are  that 1) LBE looks at the joint likelihood of the XYZ model, whereas MC only explicitly models the number of susceptibles and 2) the difference in the structure of the transition probability.  We use the delta method in conjunction with $\hat{\beta}$ and $\hat{\gamma}$ estimations from the Nelder-Mead optimization of the likelihood to estimate the stantard error of $\hat{\rr}$.
  The advantages of using this model are that our modelling assumptions are explicit and that the modelling process seeems to be reasonable.  Disadvantages include maximizing the liklihood may be more time consuming than other methods due to the need to approximate $X(t)$ and $Z(t)$ for many different values of $\beta$ and $\gamma$ and that if we are wrong about the generating process of the XYZ curves, then this model may be a very poor fit.

@@

We can also use the Delta method to derive variance estimates of LMA, LMAT, LL, (which in this case is equivalent to the standard error derived from the typical linear regression framework), and LBE.  We can also use the Delta method to derive an estimate for MC, where $\hat{\sigma}$ is the estimate of the variance of the $\hat{\alpha}$,
\begin{align*}
h(\alpha) &= \log \left ( \frac{1}{1-\alpha}\right ) \\
  h^\prime(\alpha) &= \alpha - 1 \\
                     V[ \hat{\rr} ] & \approx \frac{1}{T}  \left (\frac{\hat{\sigma}}{(\hat{\alpha} - 1)^2} \right ).
\end{align*}

@@

      \end{center}
      The baseline data set is generated by adding noise to the susceptible and infectious curves from deterministic SIR paths, which are shown in Figure \ref{fig:baseline-data}.  The recovered compartment is then adjusted to maintain a constant population value.  In this scenario, we see that the infection begins slowly but over the course of a year infects nearly 75\% of the population.
      \begin{figure}
        \centering
        \includegraphics[width=.8\textwidth]{images/baseline_sir.pdf}
        \caption{SIR (XYZ) curves for the deterministic model with $N=100,000$, $(\beta, \gamma) = (0.06, 0.03)$, $(X(0), Y(0), Z(0))= (99950, 50,0)$  The baseline data is generated by using this model and adding noise to the $X$ and $Y$ compartments and adjusting $Z$ at each time step to maintain a constant population.}\label{fig:baseline-data}
        \end{figure}
      This baseline data set is chosen due to its moderate $\rr$ size of 2, which indicates an outbreak of an epidemic similar to ZIK-V, EVD, and H1N1 pandemic influenza (see Table \ref{tab:r0-real-ex}).  The infection rate and recovery rate ($\beta, \gamma$) are relatively small.  The duration of the disease, under the true model, spans nearly a whole year (see Figure \ref{fig:baseline-data}) and so we use a year's worth of daily observations of ($\hat{X},\hat{Y}, \hat{Z}$) values.  Initially, 50 (0.05\%) infectious individuals are introduced to the population of 100,000 individuals (approximately the average population size of a US county in 2015).  The standard deviation of the $\hat{X}$ values is 0.1\% of the magnitude of the initial number of susceptible individuals, and the standard deviation of the $\hat{Y}$ values is 10\% of the magnitude of the initial number of infectious individuals, although in this setting, the number of infectious individuals increases rapidly.  We use Gaussian error as our baseline error because of its well known statistical properties so we can more easily compare methods to one another.  Later on, we explore the autoregressive and PAVA simulated data, which may better simulate data from  a real epidemic.

@@

\section{Results}\label{sec:results-top}
\subsection{$\rr$ estimation methods}\label{sec:results}
There are numerous aspects to consider when using a method to estimate $\rr$.  Among these include values of both the infection and recovery rate $(\beta, \gamma)$, the number of data points available ($T$), the total size of the population ($N$), the initial percent of susceptible individuals $\left (\frac{X(0)}{N}\right)$, the initial percent of infectious individuals $\left (\frac{Y(0)}{N}\right )$, the magnitude of the variance of both the number of susceptibles and the number of infectious ($\sigma_X^2, \sigma_Y^2$), the error modeling assumptions, data generated non-SIR models, and finally methods of estimating the variance.  For the sake of brevity, we do not provide the full results here, and instead examine the results at a high-level.  However, for those interested readers, we provide results in the form of forest plots and tables of estimates of $\rr$ for each data set in the Supplementary Material.

We implement the \wxxsir methods to estimate $\rr$ using the baseline data set with normal error.  The results are displayed in Figure \ref{fig:baseline-res} and Table \ref{tab:baseline-res}.
\begin{figure}[H]
  \centering
  \includegraphics[scale=0.5]{images/BaseBase.jpg}
  \caption{Forest plots of estimates of $\rr$ for the \xxsir methods for the baseline data set $(\beta=.06, \gamma=.03, O=365, X(0)=99950, Y(0)=50, Z(0)=0, \sigma_X=100, \sigma_Y=5, N=10^5)$.  The points are the estimates of $\rr$ and the lines denote $\pm 2\cdot $Std. Err.  The vertical, dashed line is the true $\rr$ value.}\label{fig:baseline-res}
  \end{figure}

\begin{table}[H]	
	\centering
	\begin{tabular}[t]{l|r|r}
		\hline
		Model & Estimate & Std. Err\\
		\hline
		RE & 2.0000 & 0.0056\\
		\hline
		rRE & 1.9997 & 0.0050\\
		\hline
		LMA & 1.6864 & 0.1886\\
		\hline
		LMAT & 1.4063 & 0.6309\\
		\hline
		IPR & 4.4204 & 12.3593\\
		\hline
		SIPR & 1.8871 & $<$ 1e-04 \\
		\hline
		LL & 1.9998 & 0.0002\\
		\hline
		MC & 0.9486 &  $<$ 1e-04 \\
		\hline
		SB & 1.6495 & 0.0672\\
		\hline
		LBE & 14.6784 & 1.2330\\
		\hline
	\end{tabular}
        \caption{Baseline Data $(\beta=.06, \gamma=.03, T=365, X(0)=99950, Y(0)=50, Z(0)=0, \sigma_X=100, \sigma_Y=5, N=10^5)$, $\rr$ Estimates and Standard Errors.  \com{Table does not match graph --LBE}}\label{tab:baseline-res}
\end{table}

Overall, in Figure \ref{fig:baseline-res}, we see a wide variation in the estimates of and the standard errors for $\rr$ depending on the method used.  The methods of RE, rRE, LMA, LMAT, IPR, and LL all result in a CI that covers the true value of $\rr=2$.  Of these, LMAT and IPR result in a CI which contains 1, so we would not be able to say whether or the outbreak of a disease was due to random chance or not. On the other hand, many methods appear to underestimate the true $\rr$, and many of them have larger standard errors as well.  We would say that IPR's CI is uninformative. Additionally, MC has a standard error of less than $.0001$, which is very small.  LBE, which assumes that the X and Z compartments are monotonic, overestimates $\rr$

@@

On the other hand, LBE also tends to have larger CIs and does not often capture the true value of $\rr$ when the the X and Z compartments are not monotonic functions.  This method tends to overestimate the value of $\rr$.

IPR often produces reasonable point estimates for $\rr$ but simply has too large of a CI to be considered a useful estimate.    Additionally, LMA and LMAT are generally prone to underestimating $\rr$ and have large and generally uninformative CIs.

@@

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.5]{images/start_arm.jpg}
	\caption{Forest plots of estimates of $\rr$ for the \xxsir methods for the Inits1-4 data sets $(\beta=.06, \gamma=.03, T=365,  Z(0)=0, \sigma_X=100, \sigma_Y=5, N=10^5)$.  The points are the estimates of $\rr$ and the lines denote $\pm 2\cdot $Std. Err.  The vertical line is the true $\rr$ value.}\label{fig:ar-small-I}
\end{figure}

One interesting feature we see in the results is that RE and rRE can result in very different estimations and CIs, which may seem unintuitive since rRE is a simple reparametrization of RE.  For example, in Figure \ref{fig:ar-r0}, we see that when the true value of $\rr$ is 1, rRE has a much smaller CI than RE, which may be one reason to prefer rRE to RE.  In this figure, we see when $\gamma << \beta$ (e.g. $\rr=60)$, then all estimates for $\hat{\rr}$ fail.  However, when $\beta >> \gamma$, we the estimates (besides IPR) can accurately obtain the true estimate. This figure also helps to show that RE and rRE are robust to extreme values of $\beta$ and $\gamma$ and hence extreme values of $\rr$.  \com{(For R0=1.5, why do RE, rRE, and LL look like 2 and not 1.5?))}

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.5]{images/parchange_ar.jpg}
	\caption{Forest plots of estimates of $\rr$ for the \xxsir methods for the infrec1-4 data sets $(T=365, X(0)=99950, Y(0)=50, Z(0)=0, \sigma_X=100, \sigma_Y=5, N=10^5)$.  The points are the estimates of $\rr$ and the lines denote $\pm 2\cdot $Std. Err.  The vertical line is the true $\rr$ value.}\label{fig:ar-r0}
\end{figure}

@@

%% model misspecification is terribly unrobust
We also examine circumstances where the data is generated from underlying models different from the SIR model.  The results are plotted in Figure \ref{fig:mod-rob}  All three sets of data sets generated by the non-SIR models result in poor estimates of $\rr$, meaning that even our best estimation techniques fail under model misspecification.  Moreover, these methods seem to underestimate $\rr$ \com{Illegible axes}

@@

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.5]{images/timediff.jpg}
	\caption{Forest plots of estimates of $\rr$ for the \xxsir methods for the baseline data sets with normal error structure $(\beta=.06, \gamma=.03, X(0) = 99950, Y(0)=50, \sigma_X=100, \sigma_Y=5, N=10^5)$.  The points are the estimates of $\rr$ and the lines denote $\pm 2\cdot $Std. Err.  The vertical line is the true $\rr$ value.}\label{fig:timediff}
      \end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.5]{images/obs.jpg}
	\caption{Forest plots of estimates of $\rr$ for the \xxsir methods for the baseline data sets with normal error structure $(\beta=.06, \gamma=.03, X(0) = 99950, Y(0)=50, \sigma_X=100, \sigma_Y=5, N=10^5)$.  The points are the estimates of $\rr$ and the lines denote $\pm 2\cdot $Std. Err.  The vertical line is the true $\rr$ value.}\label{fig:obs}
\end{figure}

@@

Table \ref{tab:rep-samp} shows the mean and standard deviation of $\rr$ estimates produced by the models under 100 repetitions of the baseline conditions, with new noise generated each time. We want to compare the standard deviations of the estimates to the standard errors returned by each of the models for a single data set to asses how reliable those standard errors are. Overall, it appears than most of the models have standard error estimates that are larger than the actual standard deviation of the estimates under repeated sampling. In some cases, such as LL, LMA, and LMAT, the standard deviation and standard error estimates are fairly close. The least squares methods, RE and rRE, on the other hand, tend to overestimate the sampling error by an order of magnitude. The methods SB and LBE have individual standard error estimates that are smaller than the standard deviation of repeated estimates, meaning that they understate how much error is in the estimates.

@@

Table \ref{tab:bb-samp} shows the mean and standard deviation of $\rr$ estimates produced by the models under 100 block bootstrap repetitions of the baseline data. The data were first detrended using the predictions of the compartment sizes at each time point from each of the models on the original data set. The block bootstrap algorithm was then applied on to the residuals from the predictions using a fixed block length. Only 5 of the models provide predictions of the size of the compartments and thus allow for detrending of the data. The ratio estimator and reparameterized model have fairly accurate estimates of $\rr$ as well as very small standard errors. On the other hand, LMA, LMAT, and LBE  have very inaccurate estimates as well as very large standard errors of $\rr$.

@@

Figure \ref{fig:coverage} shows the proportion of repetitions for both repeated sampling and block bootstrapping in which the 95\% confidence intervals of the estimates produced by the models covers the true $\rr$. The methods RE, rRE, LMAT, and IPR with all time points contain the true $\rr$ value in all of the confidence intervals. On the other hand, MC and SB models never cover the true $\rr$ in any of the confidence intervals.  Both LL and LBE cover the true value about 80\% of the time.

\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.5]{images/coverage2.jpg}
		\caption{Proportion of coverage of true $\rr$ of confidence intervals produced by repeated sampling for the applicable method (delta method or posterior).}
		\label{fig:coverage}	
	\end{center}
\end{figure}

% \begin{figure}[H]
% 	\begin{center}
% 		\includegraphics[scale=0.5]{images/coverage.jpg}
% 		\caption{Proportion of coverage of true $\rr$ of confidence intervals produced by repeated sampling for block bootstrapping.}
% 		\label{fig:coverage}	
% 	\end{center}
% \end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{images/h1n1.jpg}
	\caption{Results from SIR models on H1N1 data.}
\end{figure}

@@

The results of this analysis are that most of our methods yield point estimates of $\hat{\rr}$ of about 1.2, except for SB and IPR.  The methods RE, rRE, LMA, LMAT, LL, and MC all have small CIs as well.  These estimates show that one, H1N1 was indeed an epidemic in the time period but are also smaller compared to the Towers and Feng estimate.  As Towers and Feng used a different method of estimating $\rr$ than the above, this shows the importance of testing modelling assumptions when using on epidemic data.

\com{Graph to show how R0 changes with T.  X axis is time start, y axis is time end of data used.  size is magnitude of r0}

@@

%% what is r0 and why it is important
In this paper, we examine $\rr$, the average number of secondary infections a primary individual generates when introduced into a completely susceptible population, in the context of the SIR model.  The quantity $\rr$ summarizes an epidemic into a single estimate in the hopes that infectious disease outbreaks may be compared to one another.  A value of $\rr > 1$ indicates an outbreak of a disease whereas a value of $\rr < 1$ indicates that no outbreak will occur.  It is critical to not only estimate the value of $\rr$ but also a corresponding CI.  Overestimating $\rr > 1$ during an ongoing epidemic may result in a waste of resources in prevention and control measures; underestimating $\rr < 1$ may result in a worldwide crisis.

In Table \ref{tab:r0-real-ex}, we provide estimates for $\rr$ that have appeared over the past three decades for HIV, Zika, Ebola, influenza, H1N1 influenza, and measles in different locations throughout the world.  Each of these estimates is greater than $1$, which makes sense as we tend observe prolonged outbreaks of a disease as opposed to isolated occurences.  The largest $\rr$ point estimate we have seen is for measles at $\rr=17$ in England in the 20th century, and the smallest is $\rr=1.08$ for HIV in Portugal in the 21st century.
%% r0 as a property of the model
The quantity $\rr$ may summarize an epidemic, accounting for temperature, seasonality, location, and genetics of a disease, but it is still a property of the model.  Namely, how individuals progress through an epidemic (e.g. susceptible to infectious to recovered compartments) influences how we view $\rr$.  We thus should compare estimates for $\rr$ for a fixed epidemic model.  Here, we limit our analysis to estimates of $\rr$ derived from the SIR model.  Then $\rr = \frac{\beta}{\gamma}$ where $\beta$ is the rate of infection and $\gamma$ is the rate of recovery.  As a consequence, we need to estimate $\rr$ from not only the number of infections but also the number of individuals recovering at each time step.
